Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging

### 2013-03-14 HRUIonGPU/jrnl_article.tex

\author{J. P. \AA{}sen, J. I. Buskenes, C.-I. C. Nilsen, A. Austeng and S. Holm%\IEEEmembership{Member,~IEEE,}% <-this % stops a space
%\thanks{M. Shell is with the Department
%of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
%GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
%\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
%\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}
}

@@

% The paper headers
\markboth{Transactions on Ultrasonics and Ferroelectrics and Frequency Control,~Vol.~x, No.~y, March~2013}%
{\AA{}sen \MakeLowercase{\textit{et al.}}: Capon Adaptive Beamforming on the GPU}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

@@

Previous work on parallel implementations of the Capon beamformer have been focused on radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}. For medical ultrasound, Chen et al. \cite{Chen2011a, Chen2011, Chen} implemented a Capon beamformer on the GPU that produced real apodization weights based on real data. However, the Capon beamformer is meant to work on complex data. Under the constraint of real data, the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached. Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions in an ultrasound image (Fig 9. in \cite{Synnevag2009}). Complex arithmetic also requires more operations per sample, but the sampling frequency can also be reduced. The number of samples needed per image is therefore minimized. All results in this paper are reported with the use of baseband (or I/Q) data, where the sampling frequency is close to the system bandwidth.

@@

% denne må utelates i IUS-artikkelen
In recent work \cite{asen2012, Buskenes} we have applied a GPU-based Capon beamformer for both active sonar and for cardiac ultrasound imaging. Sonar has a less strict real-time requirement than medical ultrasound imaging, which means that most array configurations can be handled in real time using the GPU. Medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates at a level where even a high-end GPU can not do the processing required for a straightforward implementation of the Capon beamformer. Since the execution time is cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays. Various approaches to low-complexity Capon beamforming have therefore recently been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}. In this paper we will make use of a beamspace transformation described by Nilsen et al. \cite{Nilsen2009}, in order to maintain high framerates as the number of elements increases. The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)

@@

The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS). Here, the signal recorded at element number $m \in [0,M-1]$ at time $n$ is delayed with an appropriate delay $\Delta_m[n]$ to focus and steer the ultrasound beam. This delayed data is here denoted $x_m[n]$. The beamformer output, $z[n]$, is then the sum of all $M$ elements weighted with a factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n] = \vec{w}^H\vec{x}[n]. \label{eq:z}
\end{align}
The weight vector, or window, $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR). Although different windows can be used in different areas of the images, $\vec{w}$ is usually predefined.

@@

The Capon beamformer, following the approach in \cite{Synnevag2009}, can be divided into three main steps: estimation of a sample covariance matrix, solving a system of linear equation, and calculation of the beamformer output.  The sample covariance matrix is estimated as 
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
\end{align}
where  $N_K = 2K + 1$ is the number of samples included in time, $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l^\text{th}$ subarray of length $L$, and $N_L = M-L+1$ is the number of subarrays. Note that the covariance matrix dimension is $L \times L$. The covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness, 
\begin{align}\label{eq:diag}
\mat{\hat{R}}[n] =  f(\vec{x}[n]) = \mat{\breve{R}}[n] + \epsilon[n]\mat{I}.
\end{align}
A weight proportional to the output power is often used to reduce the need for parameter adjustments, and the trace of $\mat{\breve{R}}$, 
\begin{align}\label{eq:diag_adapt}
\epsilon[n] &= d \times \frac{\text{trace}\{\mat{\breve{R}}[n]\}}{L},
\end{align}
has been applied in much of the recent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}. It has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.

@@

In order to get a well-conditioned $\mat{\hat{R}}$, avoid signal cancellation, and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim T_p/T_s$ time samples \cite{Synnevag2007a}, where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period. For our baseband data, one pulse length ($1.5\lambda$) is around 3 samples ($K=1$).

@@

In this section we give a brief discussion of GPU computing, laying the basis for further discussion on parallel implementation strategies. As GPU framework, we have selected CUDA by Nvidia \cite{Nvidia2013}, however the challenges and proposed methods described in this paper are valid for other GPU framworks as well, like e.g. OpenCL. The key point is that GPUs can accelerate large sets of small concurrent tasks, like beamforming a pixel in a large image, much better than a CPU. While the CPU has few cores and little logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied.

The CUDA compute model is based on the execution of a kernel function over a large set of threads, where the kernel function describes the work to be done in each thread. As depicted in Fig. \ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block ($32\times32$). Threads inside a block support low-cost synchronization if needed. From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of 32 or more concurrent multiplications and/or additions, depending on the architecture. A group of blocks is scheduled to each SM until all blocks have been processed.

An SM has a limited amount of near-core, or local, memory and registers. Hence, to achieve maximum GPU occupancy (having the maximum number of threads residing on each SM), each thread has to be as light on register and memory consumption as possible. It is also important that each thread performs enough instructions per transferred byte in order to hide memory latency, and that the problem can be divided into enough threads to hide instruction latency. Reads and writes to and from global GPU memory, in addition to CPU-GPU transfers, are slow and should be minimized as much as possible. All of these constraints combined pose a real challenge as the target problem has to map well to the GPU framework on a grid, block, thread, and memory consumption level in order to benefit from parallel acceleration. In the next section, we will describe how this mapping can be done for the Capon beamforming algorithm.

\begin{figure}
\centerline{\includegraphics[width=3.5in]{gfx/gpu_layout_vertical.eps}}
\caption{Calculation of Capon adaptive weights mapped to GPU architecture. A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups. Note that several samples can be processed per block of threads if $L$ is small. The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
\label{fig:gpulayout}
\end{figure}

@@

Calculation of one element of \R is from (\ref{eq:R}) found to be independent of calculation of other elements. A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ \cite{Chen2011}. This will minimize global reads and writes needed per thread, but since we want to avoid block-to-block synchronization, the maximum subarray length is limited to 32 by this approach. Since most medical ultrasound arrays have more than 32 elements, an implementation not limited by $L_{max}=32$ is preferable.

To reach this goal we start by showing how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays and in time. Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
\end{align}
In time we have that
\begin{align}
\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
Both sliding across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ \cite{Buskenes}. However, since $N_K$ typically is small, and the GPU has a limited amount of memory available per thread, we focus on sliding across subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_K(L(N_L-2)+2L^2))$. For $L=M/2$ this reduces to $O(N_KM^2)$. Note that sliding will break independence along the sliding dimension, thus, calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$. One thread is therefore used per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{fig:gpulayout}). Maximum supported $L$ is increased to the maximum number of threads per block (currently 1024) with this approach. However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full occupancy of the GPU.

@@

\subsection{Solving Multiple Small Linear Systems}
The literature has mostly focused on large systems when solving systems of linear equations on the GPU. GPU libraries therefore often lack a comprehensive collection of batched solvers for large sets of small systems. The reason for this is two-fold. First the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU. Second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available. Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared to the CPU in every case. It is therefore not given that solving on the GPU is faster than solving on the CPU in general. However, there are two arguments for solving on the GPU. First the CPU, which might already run a lot of ultrasound-related algorithms, is offloaded. Second, the transfer of several thousand covariance matrices over the PCI-Express bus, from the CPU to the GPU, is avoided.

@@

\subsection{Compute Beamformer Output}
To calculate the beamformer output, equation (\ref{eq:z_mv}) has been implemented using one thread per element weight ($L$ threads per weight vector). After the solution vector $\vec{b}$ has been placed in near-core memory, one of the $L$ threads finds the sum of $\vec{b}$. 
%Another option is to let several threads cooperate to form the sum using a binary reduction sum. However, in the following benchmarks the first approach has been used. 
Each of the $L$ threads then reads one element of $\vec{b}$, and scales it with the inverse sum to get the weight vector $\vec{w}[n]$.

@@

\section{Beamspace Processing}\label{sec:bs}
Nilsen et al. \cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations of the Capon beamformer when applied to medical ultrasound imaging. Beamspace is an alternative representation of the per-element data, transformed into a fan of beams covering the sector illuminated by transmit beams. The transformation is formally stated as
\begin{align}
\vec{x}_\text{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \ \ \ \ \ \ \ \ \ \ \, \mat{B} \in \mathbb{C}^{N,M}\\
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M} \ \ \ N \le M.
\end{align} 
The steering vectors in $\mat{B}$, the so-called Butler matrix, determine each beam's direction in space. Beams where no interference is present can therefore be removed by simply removing rows from $\mat{B}$.

For a focused system like cardiac ultrasound imaging, where the received signal is concentrated in a narrow sector around the transmit direction, a small percentage of beams contains almost all received energy. Nilsen et al. concluded that as little as three beams around (and including) the transmit direction were adequate to produce results comparable to applying Capon beamforming in element space. Hence, for $\lambda$-factor spaced arrays the inversion steps is found to be constant with respect to the number of array elements.

The beamspace version of the Capon beamformer is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix 
\begin{align}
\mat{\hat{R}}_\text{BS} = f(\vec{x}_\text{BS}) = f(\mat{B}\vec{x}), 
\end{align}
where $\mat{B} \in \mathbb{C}^{N_b,L}$, and $N_b$ is the number of selected beams. If the transformation is applied before the covariance matrix is constructed, one transformation is performed per subarray, yielding subarrays that no longer overlap in space. The amount of input data are then changed from $M$ to $N_LN_b$ elements per sample, however all remaining steps will get lower computational complexity when $N_b < L$.

Another option is to transform the element space covariance matrix to beamspace. 
\begin{align}
\mat{\hat{R}}_\text{BS} &= \mat{B}\mat{\hat{R}}\mat{B}^H.
\end{align}
The advantage of this approach is that minor changes have to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space. In this paper we will focus on the first approach, and how it has been implemented on the GPU.
%By this approach we go from solving $\mat{\hat{R}}\vec{b} = \vec{1}_L$, which is an $L \times L$ system, to solving the $N_b \times N_b$ system $\mat{\hat{R}}_{BS}\vec{b}_{BS} = (\sqrt{L}\vec{e}_1)$. Now, one can either choose to transform the subarray sums, as in (\ref{eq:subsum}), to beamspace or transform the result, $\vec{b}_{BS}$, back to element space. 
In both cases solving has to be performed against a beamspace steering vector
\begin{align}
\vec{a}_\text{BS} &= \mat{B}\vec{1}_L = \sqrt{L}\vec{e}_1,
\end{align}
where $\vec{e}_i$ has the value one in the i$^{th}$ position and zeroes elsewhere.

%The implementation calculating covariance matrices therefore needs to be adjusted to handle non-overlapping subarrays. The complexity of constructing the covariance matrix is changed to $O(N_b^2L)$, so constructing the covariance matrices will be faster in beamspace as long as $N_b < \sqrt{L}$.
%Derivation of sliding beamspace transformation. 
Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l^{th}$ subarray for the $k^{th}$ beam is:
\begin{align}\label{eq:bs_formula}
x_{\text{BS},l,k} = \frac{1}{\sqrt{L}}\sum_{p=i}^{l+L-1}x_p e^{-j2\pi k(p-l)/L} , k \in [0, N_b-1].
\end{align}
Manipulating this expression we find that for $l \in [1, L-1]$
\begin{align}\label{eq:sliding_bs}
x_{\text{BS},l,k} &= \frac{1}{\sqrt{L}}((x_{\text{BS},l-1,k} - x_{l-1})e^{-j2\pi k/L} \nonumber \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \nonumber \\
&= \frac{1}{\sqrt{L}}(x_{\text{BS},l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}.
\end{align}
We denote this the sliding beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}. For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).

@@

\begin{figure}
%\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_32_40_400.png} \label{fig:benchUS32}}
%\hfil
%\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_606.png}%
%\label{fig:benchUS64}}}
\centerline{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_540.png}}
\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a $70\degree$ wide $15$ cm deep sector. Execution time is plotted as a function of subarray length $L$. The plot shows execution times for the three steps listed in Section \ref{sec:meth}. Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.}
\label{fig:bench}
\end{figure}

>>>>

\author{J. P. \AA{}sen, \IEEEmembership{Student Member,~IEEE,} J. I. Buskenes, C.-I. C. Nilsen, \IEEEmembership{Member,~IEEE,} A. Austeng, \IEEEmembership{Member,~IEEE,} and S. Holm, \IEEEmembership{Senior Member,~IEEE,}% <-this % stops a space
\thanks{Manuscript received ...; accepted ...}
\thanks{Jon Petter \AA{}sen and Sverre Holm is with the Medical Imaging Lab (MI-Lab) at the Norwegian University of Sience and Technology, Trondheim, Norway. Jo Inge Buskenes, Carl-Inge Colombo Nilsen, Andreas Austeng and Sverre Holm is with the Department of Informatics, University of Oslo, Oslo, Norway. (e-mail: jon.p.asen@ntnu.no)} 
}

@@

% The paper headers
\markboth{Transactions on Ultrasonics and Ferroelectrics and Frequency Control,~Vol.~x, No.~y, Z~201\ae{}}%
{\AA{}sen \MakeLowercase{\textit{et al.}}: Capon Adaptive Beamforming on the GPU}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

@@

Previous work on parallel implementations of the Capon beamformer have been focused on radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}. For medical ultrasound, Chen et al. \cite{Chen2011a, Chen2011, Chen} implemented a Capon beamformer on the GPU that produced real apodization weights based on real data. However, the Capon beamformer is meant to work on complex data. Under the constraint of real data, the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached. Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions in an ultrasound image \cite[Fig 9.]{Synnevag2009}. Complex arithmetic also requires more operations per sample, but the sampling frequency can also be reduced. The number of samples needed per image is therefore minimized. All results in this paper are reported with the use of baseband (or I/Q) data, where the sampling frequency is close to the system bandwidth.

@@

% denne må utelates i IUS-artikkelen
In recent work \cite{asen2012, Buskenes} we have applied a GPU-based Capon beamformer for both active sonar and for cardiac ultrasound imaging. Sonar has a less strict real-time requirement than medical ultrasound imaging, which means that most array configurations can be handled in real time using the GPU. Medical ultrasound imaging, and in particular cardiac ultrasound imaging, demands frame rates at a level where even a high-end GPU can not do the processing required for a straightforward implementation of the Capon beamformer. Since the execution time is cubic with respect to the number of array elements, we need methods that reduce the computational complexity for large arrays. Various approaches to low-complexity Capon beamforming have therefore recently been proposed in the literature \cite{Synnevag2011, Asl2012, Jensen2012, Kim}. In this paper we will make use of a beamspace transformation described by Nilsen and Hafizovic \cite{Nilsen2009}, in order to maintain high framerates as the number of elements increases. The beamspace transformation matches well with the focused transmit beams used in cardiac ultrasound imaging. 
%(Referer til JOE-artikkelen, få med beskrivelse av hva som blir anerledes for ultralyd.)

@@

The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS). Here, the signal recorded at element number $m \in [0,M-1]$ at time $n$ is delayed with an appropriate delay $\Delta_m[n]$ to focus and steer the ultrasound beam. This delayed data is here denoted $x_m[n]$. The beamformer output $z[n]$ is then the sum of all $M$ elements weighted with a factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n] = \vec{w}^H\vec{x}[n]. \label{eq:z}
\end{align}
The weight vector, or window, $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR). Although different windows can be used in different areas of the images, $\vec{w}$ is usually predefined.

@@

The Capon beamformer, following the approach in \cite{Synnevag2009}, can be divided into three main steps: estimation of a sample covariance matrix, solving a system of linear equation, and calculation of the beamformer output.  The sample covariance matrix is estimated as 
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
\end{align}
where  $N_K = 2K + 1$ is the number of samples included in time, $\vec{x}_l = [x_l[n], \dotso, x_{l+L-1}[n]]$ is the $l\text{th}$ subarray of length $L$, and $N_L = M-L+1$ is the number of subarrays. Note that the covariance matrix dimension is $L \times L$. The covariance matrix is loaded with a diagonal factor $\epsilon$ to ensure numerical stability and increased robustness, 
\begin{align}\label{eq:diag}
\mat{\hat{R}}[n] = \mat{\breve{R}}[n] + \epsilon[n]\mat{I} = f(\vec{x}[n]).
\end{align}
Where $f(\cdot)$ represents the process of constructing $\mat{\hat{R}}$ from the input $\vec{x}$.
A factor proportional to the output power is often used to reduce the need for parameter adjustments, and the trace of $\mat{\breve{R}}$, 
\begin{align}\label{eq:diag_adapt}
\epsilon[n] &= d \times \frac{\text{trace}\{\mat{\breve{R}}[n]\}}{L},
\end{align}
has been applied in much of the recent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Nilsen2009, Wang2009, Mehdizadeh2012}. It has also been applied to HF and VHF antenna processing \cite{Featherstone1997b}.

@@

In order to get a well-conditioned $\mat{\hat{R}}$, avoid signal cancellation, and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N_K \sim T_p/T_s$ time samples \cite{Synnevag2007a}, where, $T_p$ is the pulse length in seconds and $T_s$ is the sampling period. For our baseband data, one pulse length ($1.5\lambda$) is around 3 samples ($K=1$). Typical values for $d$ is between $1/10$ and $1/100$.

@@

In this section we give a brief discussion of GPU computing, laying the basis for further discussion on parallel implementation strategies. As GPU framework, we have selected CUDA by Nvidia \cite{Nvidia2013}, however the challenges and proposed methods described in this paper are valid for other GPU framworks as well, like e.g. OpenCL. The key point is that GPUs can accelerate large sets of small concurrent tasks, like beamforming a pixel in a large image, much better than a CPU. While the CPU has few cores and little logic to process complex threads, the GPU has many cores, but each core needs many (preferably simple) threads to keep it occupied and to achieve maximum throughput.

The CUDA compute model is based on the execution of a kernel function over a large set of threads, where the kernel function describes the work to be done in each thread. As depicted in Fig. \ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block (e.g. $32\times32$). Threads inside a block support low-cost synchronization if needed. From a CUDA perspective the GPU consist of one or more streaming multi-processors (SM), where each SM is capable of 32 or more concurrent multiplications and/or additions, depending on the architecture. A group of blocks is scheduled to each SM until all blocks have been processed.

An SM has a limited amount of local (near-core or shared) memory and registers. Therefore, to achieve maximum GPU occupancy (defined as having the maximum number of threads residing on each SM), each thread has to consume as few registers and as little local memory as possible. Each thread should also perform enough instructions per transferred byte in order to hide memory latency. Dividing the problem into enough threads will help to hide instruction latency. Global GPU memory transactions, in addition to CPU-GPU transfers, are slow and must be minimized. All of these constraints combined pose a real challenge as the target problem has to map well to the GPU framework on a grid, block, thread, and memory consumption level in order to benefit from parallel acceleration. In the next section, we will describe how this mapping can be done for the Capon beamforming algorithm.

\begin{figure}
\centerline{\includegraphics[width=3.5in]{gfx/gpu_layout_vertical.eps}}
\caption{Calculation of Capon adaptive weights mapped to GPU architecture. A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups. Note that several samples can be processed per thread-block if $L$ is small. The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
\label{fig:gpulayout}
\end{figure}

@@

Calculation of one element of \R is from (\ref{eq:R}) found to be independent of calculation of other elements. A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ \cite{Chen2011}. This will minimize global reads and writes needed per thread, but since we want to avoid block-to-block synchronization, the maximum subarray length is limited to 32 by this approach. Since most medical ultrasound arrays have 64 elements or more, an implementation not limited by $L_{max}=M/2=32$ is preferable.

To reach this goal we start by showing how the cubic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that calculations in (\ref{eq:R}) overlap both across subarrays and in time. Element $(i,j)$ of $\mat{\breve{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ in a sliding window maner as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  \nonumber \\
&+ \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}[n']x_{j+N_L-1}[n']^* \nonumber \\
&- x_{i-1}[n']x_{j-1}[n']^H). \label{eq:sliding}
\end{align}
In time we have that
\begin{align}
\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}_l[n+K]\vec{x}_l[n+K]^H \nonumber \\
&- \vec{x}[n-K-1]\vec{x}[n-K-1]^H).
\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
Sliding windows across subarrays and in time can be realized simultaneously if $\mat{\hat{R}}$, averaged over subarrays, is calculated from an $N_K$-average in time of the outer products $\vec{x}[n]\vec{x}[n]^H$ \cite{Buskenes}. However, since $N_K$ typically is small, and the GPU has a limited amount of local memory available per thread, we focus on subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(N_K(L(N_L-2)+2L^2))$. For $L=M/2$ this reduces to $O(N_KM^2)$. The sliding-window approach will cause dependence among the iterative calculations. The level of task granularity is thereby increased. When sliding across subarrays, calculations are still independent across time, but not along diagonals of $\mat{\hat{R}}$. One thread is therefore used per diagonal in the upper triangle of $\mat{\hat{R}}$, and for small values of $L$, several covariance matrices are calculated per thread block (Fig. \ref{fig:gpulayout}). Maximum supported $L$ is now increased to the maximum number of threads per block (currently 1024). However, for small-$L$-high-$M$ configurations, memory consumption per thread will be too high to achieve full occupancy of the GPU.

@@

\subsection{Solving Multiple Small Linear Systems}
The literature has mostly focused on large systems when solving systems of linear equations on the GPU. GPU libraries therefore often lack a comprehensive collection of batched solvers for large sets of small systems. The reason for this is two-fold. First the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU. Second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of local memory and registers are available. Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared to the CPU in every case. It is therefore not given that solving on the GPU is faster than solving on the CPU in general. However, there are two arguments for solving on the GPU. First the CPU, which might already run a lot of ultrasound-related algorithms, is offloaded. Second, the transfer of several thousand covariance matrices over the PCI-Express bus, from the CPU to the GPU, is avoided.

@@

\subsection{Compute Beamformer Output}
To calculate the beamformer output, (\ref{eq:z_mv}) has been implemented using one thread per element weight ($L$ threads per weight vector). After the solution vector $\vec{b}$ has been placed in local memory, one of the $L$ threads finds the sum of $\vec{b}$. 
%Another option is to let several threads cooperate to form the sum using a binary reduction sum. However, in the following benchmarks the first approach has been used. 
Each of the $L$ threads then reads one element of $\vec{b}$, and scales it with the inverse sum to get the weight vector $\vec{w}[n]$.

@@

\section{Beamspace Processing}\label{sec:bs}
Nilsen and Hafizovic \cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations of the Capon beamformer when applied to medical ultrasound imaging. Beamspace is an alternative representation of the per-element data, transformed into a fan of beams covering the sector illuminated by transmit beams. The transformation is formally stated as
\begin{align}
\vec{x}_\text{BS}[n] &= \mat{B}\vec{x}[n], \ \ \ \ \ \ \ \ \ \ \ \ \ \, \mat{B} \in \mathbb{C}^{N,M}\\
[\mat{B}]_{p,q} &= \frac{1}{\sqrt{M}}e^{-j 2 \pi p q / M} \ \ \ N \le M.
\end{align} 
The steering vectors in $\mat{B}$, the so-called Butler or normalized DFT matrix, determine each beam's direction in space. Beams where no interference is present can therefore be removed by simply removing rows from $\mat{B}$.

For a focused system like cardiac ultrasound imaging, where the received signal is concentrated in a narrow sector around the transmit direction, a small percentage of beams contains almost all received energy. Nilsen et al. concluded that as little as three beams around (and including) the transmit direction were adequate to produce results comparable to applying Capon beamforming in element space (Section \ref{sec:meth}). Hence, for  equally pitched arrays the inversion steps is found to be constant with respect to the number of array elements.

The beamspace version of the Capon beamformer is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix 
\begin{align}
\mat{\hat{R}}_\text{BS} = f(\vec{x}_\text{BS}) = f(\mat{B}\vec{x}), 
\end{align}
where $\mat{B} \in \mathbb{C}^{N_b,L}$, $N_b$ is the number of selected beams, and $f(\cdot)$ is as defined in (\ref{eq:diag}). If the transformation is applied before the covariance matrix is constructed, one transformation is performed per subarray, yielding subarrays that no longer overlap in space. The amount of input data are then changed from $M$ to $N_LN_b$ elements per sample, however all remaining steps will get lower computational complexity when $N_b < L$.

Another option is to transform the element space covariance matrix to beamspace. 
\begin{align}
\mat{\hat{R}}_\text{BS} &= \mat{B}\mat{\hat{R}}\mat{B}^H.
\end{align}
The advantage of this approach is that minor changes have to be made to the pipeline described in Section \ref{sec:meth}, but it will only speed up the solving step if the solution $\vec{b}_{BS}$ is transformed back to element space. In this paper we will focus on the first approach, and how it has been implemented on the GPU.
%By this approach we go from solving $\mat{\hat{R}}\vec{b} = \vec{1}_L$, which is an $L \times L$ system, to solving the $N_b \times N_b$ system $\mat{\hat{R}}_{BS}\vec{b}_{BS} = (\sqrt{L}\vec{e}_1)$. Now, one can either choose to transform the subarray sums, as in (\ref{eq:subsum}), to beamspace or transform the result, $\vec{b}_{BS}$, back to element space. 
In both cases solving has to be performed against a beamspace steering vector
\begin{align}
\vec{a}_\text{BS} &= \mat{B}\vec{1}_L = \sqrt{L}\vec{e}_1,
\end{align}
where $\vec{e}_i$ has the value one in the $i$th position and zeroes elsewhere.

%The implementation calculating covariance matrices therefore needs to be adjusted to handle non-overlapping subarrays. The complexity of constructing the covariance matrix is changed to $O(N_b^2L)$, so constructing the covariance matrices will be faster in beamspace as long as $N_b < \sqrt{L}$.
%Derivation of sliding beamspace transformation. 
Given a $M$ long data vector divided into $N_L$ subarrays, the beamspace transformation of the $l$th subarray for the $k$th beam is:
\begin{align}\label{eq:bs_formula}
x_{\text{BS},l,k} = \frac{1}{\sqrt{L}}\sum_{p=i}^{l+L-1}x_p e^{-j2\pi k(p-l)/L} , k \in [0, N_b-1].
\end{align}
Manipulating this expression we find that for $l \in [1, L-1]$
\begin{align}\label{eq:sliding_bs}
x_{\text{BS},l,k} &= \frac{1}{\sqrt{L}}((x_{\text{BS},l-1,k} - x_{l-1})e^{-j2\pi k/L} \nonumber \\ &+ x_{i+L-1}e^{-j2\pi k(L-1)/L}) \nonumber \\
&= \frac{1}{\sqrt{L}}(x_{\text{BS},l-1,k} - x_{l-1} + x_{i+L-1})e^{-j2\pi k/L}.
\end{align}
We denote this the sliding-window beamspace transformation (SBS), and it is equal to a normalized sliding DFT \cite{Lyons2003}. For $l=0$ a full computation has to be performed using (\ref{eq:bs_formula}).

@@

\begin{figure}
%\centerline{\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_32_40_400.png} \label{fig:benchUS32}}
%\hfil
%\subfloat[]{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_606.png}%
%\label{fig:benchUS64}}}
\centerline{\includegraphics[width=3.5in]{gfx/benchmark_bar_64_80_540.png}}
\caption{Benchmark of GPU Capon beamforming in element space for a cardiac image covering a $70\degree$ wide $15$ cm deep sector. Execution time is plotted as a function of subarray length $L$. The plot shows execution times for the three steps listed in Section \ref{sec:meth}: Calculating covariance matrices with three different choices of temporal averaging, solver, and the final beamformer sum.}
\label{fig:bench}
\end{figure}
